---
title: "Practice with dplyr and ggplot2"
author: "Dr. Shoemaker"
date: "DATA 2401"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Fancy Histograms!

We'll start by making some histograms.

```{r}
#install.packages("dslabs")
library(dslabs) 
data(heights)
glimpse(heights)
```

This data is the heights of humans, divided by their biological sex.

Use ggplot to make a histogram of all of the heights:

```{r}
ggplot(heights, aes(x = height)) +
  geom_histogram(binwidth = 1, color = "black", fill = "lightblue") +
  labs(title = "Histogram of Human Heights",
       x = "Height (inches)",
       y = "Count")

```

Change up the binwidth and see how the plots change. Try 1, 5, 10, and 20

```{r}

# 1
ggplot(heights, aes(x = height)) +
  geom_histogram(binwidth = 1, color = "black", fill = "lightblue") +
  labs(title = "Histogram of Heights (Binwidth = 1)",
       x = "Height (inches)",
       y = "Count")

#  5
ggplot(heights, aes(x = height)) +
  geom_histogram(binwidth = 5, color = "black", fill = "lightgreen") +
  labs(title = "Histogram of Heights (Binwidth = 5)",
       x = "Height (inches)",
       y = "Count")

#  10
ggplot(heights, aes(x = height)) +
  geom_histogram(binwidth = 10, color = "black", fill = "salmon") +
  labs(title = "Histogram of Heights (Binwidth = 10)",
       x = "Height (inches)",
       y = "Count")

#  20
ggplot(heights, aes(x = height)) +
  geom_histogram(binwidth = 20, color = "black", fill = "orchid") +
  labs(title = "Histogram of Heights (Binwidth = 20)",
       x = "Height (inches)",
       y = "Count")

```

Smooth this out to an emperical density with `geom_density()`

```{r}
ggplot(heights, aes(x = height)) +
  geom_density(fill = "skyblue", alpha = 0.5) +
  labs(title = "Empirical Density of Human Heights",
       x = "Height (inches)",
       y = "Density")
```

Use a new argument in the `aes()`, `group =` to split this density by sex

```{r}
ggplot(heights, aes(x = height, group = sex, fill = sex)) +
  geom_density(alpha = 0.5) +
  labs(title = "Empirical Density of Heights by Sex",
       x = "Height (inches)",
       y = "Density")
```

OR we can do it with `color` or `fill.` If you say you want to color by sex, R knows that you want a different curve for each of them.

```{r}
#color 

ggplot(heights, aes(x = height, color = sex)) +
  geom_density() +
  labs(title = "Density of Heights by sex (Color Only)",
       x = "Height (inches)",
       y = "Density")

#fill

ggplot(heights, aes(x = height, fill = sex)) +
  geom_density(alpha = 0.5) +
  labs(title = "Density of Heights by Sex (Filled)",
       x = "Height (inches)",
       y = "Density")


```

If you've used fill, then there is now a slight issue that they are overlapped. We can fix this with alpha transparency!

```{r}
ggplot(heights, aes(x = height, fill = sex)) +
  geom_density(alpha = 0.5) +
  labs(title = "Density of Heights by Sex (With Transparency)",
       x = "Height (inches)",
       y = "Density")

```

Let's make some boxplots of the same information.

```{r}
ggplot(heights, aes(x = sex, y = height, fill = sex)) +
  geom_boxplot(alpha = 0.5) +
  labs(title = "Boxplot of Heights by Sex",
       x = "Sex",
       y = "Height (inches)")

```

## Quantatitive summaries:  

Find the mean and median overall. 

```{r}
mean_height <- mean(heights$height)
median_height <- median(heights$height)

mean_height
median_height
```

Find the mean and median for both groups. 

```{r}
heights %>%
  group_by(sex) %>%
  summarize(
    mean_height = mean(height),
    median_height = median(height),
    count = n()
  )

```

How tall is the tallest woman? How short is the shortest man? 

```{R}
# Tallest woman
heights %>%
  filter(sex == "Female") %>%
  summarize(tallest_woman = max(height))

# Shortest man
heights %>%
  filter(sex == "Male") %>%
  summarize(shortest_man = min(height))

```

## Presidental Elections Data

```{r}
# install.packages("pscl")
library(pscl) # loads in the package that has this data. 
## You might need to install this...

# data for presidental elections
votedata <-  presidentialElections
glimpse(votedata)
```

Let's look at the democratic vote by state for 2000. We can't use `geom_bar` for a bar chart, since we have the category in one variable and the "height" of the bar in another. We need `geom_col()`

Make a bar graph of the democratic vote by state in 2000. 

```{r}
votedata <- presidentialElections


data(presidentialElections)


votedata_2000 <- presidentialElections %>%
  filter(year == 2000)


ggplot(votedata_2000, aes(x = reorder(state, demVote), y = demVote)) +
  geom_col(fill = "blue", alpha = 0.7) +
  coord_flip() +
  labs(title = "Democratic Vote Share by State (2000 Election)",
       x = "State",
       y = "Democratic Vote Share (%)")


```

Well this looks awful. We have two options: swap the x and y or the more fun sounding... Coordinate flip! 

Use `coord_flip()` on the previous graph to make it better. 

```{r fig.height=7, fig.width=5}

votedata_2000 <- presidentialElections %>%
  filter(year == 2000)


ggplot(votedata_2000, aes(x = reorder(state, demVote), y = demVote)) +
  geom_col(fill = "blue", alpha = 0.7) +
  coord_flip() +
  labs(title = "Democratic Vote Share by State (2000 Election)",
       x = "State",
       y = "Democratic Vote Share (%)")

```

I don't love the squashed together coordinates, but it's a display window issue.

So. This is a helpful graph, but it would be *more* helpful if it was ordered. 
Use `x = reorder(x_variable, y_variable)` in `aes()` to order the x variable by the y variable

```{r}

ggplot(votedata_2000, aes(x = reorder(state, demVote), y = demVote)) +
  geom_col(fill = "blue", alpha = 0.7) +
  coord_flip() +
  labs(title = "Ordered Democratic Vote Share by State (2000 Election)",
       x = "State",
       y = "Democratic Vote Share (%)")


```

So, what if I want to see what the north and south states did different?

start with a facet_wrap using the `south` variable:

```{r}
ggplot(votedata_2000, aes(x = reorder(state, demVote), y = demVote)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~south) +
  labs(title = "Democratic Vote Share by State (2000) - Faceted by North vs South",
       x = "State",
       y = "Democratic Vote Share (%)")

```

Okay, that's not great. Lets *color* by south instead.

```{r}
ggplot(votedata_2000, aes(x = reorder(state, demVote), y = demVote, fill = south)) +
  geom_col() +
  coord_flip() +
  labs(title = "Democratic Vote Share by State (2000)",
       x = "State",
       y = "Democratic Vote Share (%)")

```

I'm a good data scientist, so I want my plot to have a name! and my axes to have lables! Use `labs` to add a title, subtitle, and x and y labels.

```{r, eval = F}
ggplot(votedata_2000, aes(x = reorder(state, demVote), y = demVote, fill = south)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Democratic Vote Share by State (2000)",
    subtitle = "Colored by Southern vs Non-Southern States",
    x = "State",
    y = "Democratic Vote Share (%)",
    fill = "Southern State"
  )

```

You can move the legend with `theme(legend.position = "bottom")`

```{r, eval = F}
ggplot(votedata_2000, aes(x = reorder(state, demVote), y = demVote, fill = south)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Democratic Vote Share by State (2000)",
    subtitle = "Colored by Southern vs Non-Southern States",
    x = "State",
    y = "Democratic Vote Share (%)",
    fill = "Southern State"
  ) +
  theme(legend.position = "bottom")


```

What else could we facet by? years! Let's filter to year in 2008 and 2016, then facet by years.

```{r}
votedata_subset <- presidentialElections %>%
  filter(year %in% c(2008, 2016))

ggplot(votedata_subset, aes(x = reorder(state, demVote), y = demVote, fill = south)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~year) +
  labs(
    title = "Democratic Vote Share by State (2008 vs 2016)",
    subtitle = "Faceted by Year, Colored by Southern vs Non-Southern States",
    x = "State",
    y = "Democratic Vote Share (%)",
    fill = "Southern State"
  ) +
  theme(legend.position = "bottom")


```

We need to know who won! We could add a vertical line at 50 for who got more, to indicate the majority of votes. Adding the layer `geom_hline()` adds a horizontal line. (What do you guess `geom_vline()` would do?)

```{r}
ggplot(votedata_subset, aes(x = reorder(state, demVote), y = demVote, fill = south)) +
  geom_col() +
  geom_hline(yintercept = 50, linetype = "dashed", color = "red", size = 1) +  # <--- adds the 50% line
  coord_flip() +
  facet_wrap(~year) +
  labs(
    title = "Democratic Vote Share by State (2008 vs 2016)",
    subtitle = "Dashed Red Line Indicates 50% Majority Threshold",
    x = "State",
    y = "Democratic Vote Share (%)",
    fill = "Southern State"
  ) +
  theme(legend.position = "bottom")

```

### Getting fancy with a map!

When using geom_polygon or geom_map, you will typically need two data frames:

-   one contains the coordinates of each polygon (positions)
-   the other the values associated with each polygon (values).

An id variable links the two together. 

Run the below code to get a map graph. 

```{r}
library(maps)
votedata$state <- tolower(votedata$state)  ## states need to be lowercase for linking

states_map <- map_data("state") ## this gives us the lat and long for each point of each state.
  
map_plot <-  ggplot(data =  votedata %>% filter(year == 2008), aes(map_id = state)) +
    geom_map(aes(fill = demVote), map = states_map) +
    expand_limits(x = states_map$long, y = states_map$lat)
map_plot  
```

```{r}
map_plot <-  ggplot(data =  votedata %>% filter(year == 2016), aes(map_id = state)) +
  geom_map(aes(fill = demVote), map = states_map)+
  expand_limits(x = states_map$long, y = states_map$lat)
map_plot  
```

What if I want a map that shows which of the states are "south"? What do I change?

```{r}
map_plot <- ggplot(data = votedata %>% filter(year == 2016), aes(map_id = state)) +
  geom_map(aes(fill = south), map = states_map) +
  expand_limits(x = states_map$long, y = states_map$lat)
map_plot


```


### Some more dplyr practice

I want to know the average democratic vote for N vs S, by year.

First, find the average democratic votes for the north and the south, every year. You'll need to do a *double* `group_by()` here. You do it in one call of the function.  

```{r}
votedata %>%
  group_by(south, year) %>%
  summarize(avg_dem_vote = mean(demVote), .groups = 'drop')

```

Then, let's plot that! Pipe the result of your group_by and summarize to ggplot and geom_line(), with year on the x axis and your summarized value on the y axis. Color by the `south` variable. 

```{r}
votedata %>%
  group_by(south, year) %>%
  summarize(avg_dem_vote = mean(demVote), .groups = 'drop') %>%
  ggplot(aes(x = year, y = avg_dem_vote, color = south)) +
  geom_line(size = 1.2) +
  labs(
    title = "Average Democratic Vote by Region Over Time",
    x = "Year",
    y = "Average Democratic Vote (%)",
    color = "Southern State"
  ) +
  theme_minimal()

```

## Layering plots!

Penguins!

```{r}
library(palmerpenguins)
glimpse(penguins)
```

We can use boxplots to visualize the distribution of weight (body_mass_g) within each species:

```{r}
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot() +
  labs(
    title = "Penguin Body Mass by Species",
    x = "Species",
    y = "Body Mass (g)"
  )

```

What if we also want the points? Layering!! Add a geom_point to your existing boxplot. geom_boxplot + geom_point!

```{r}
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot() +
  geom_point(alpha = 0.5) +   # alpha for slight transparency
  labs(
    title = "Penguin Body Mass by Species with Points",
    x = "Species",
    y = "Body Mass (g)"
  )


```

But, these are all stacked up... to actually see them, use "geom_jitter" instead of points

```{r}
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.5, width = 0.2) +  # jitter spreads points horizontally
  labs(
    title = "Penguin Body Mass by Species with Jittered Points",
    x = "Species",
    y = "Body Mass (g)"
  )


```

How to get the boxplots on top? The layers are plotted in the order you give them, so change to geom_point + geom_boxplot. (You might want to change the alpha on the boxplot to be able to see the plots under them)

```{r}

ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_jitter(alpha = 0.5, width = 0.2) +  # draw points first
  geom_boxplot(alpha = 0.7) +              # draw boxplots on top
  labs(
    title = "Penguin Body Mass by Species",
    subtitle = "Boxplots Overlaid on Jittered Points",
    x = "Species",
    y = "Body Mass (g)"
  )


```

Maybe let's try replacing the boxplot with a `geom_violin()`?

```{r}

ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_jitter(alpha = 0.5, width = 0.2) +  # points first
  geom_violin(alpha = 0.7) +               # violin on top
  labs(
    title = "Penguin Body Mass by Species",
    subtitle = "Violin Plot with Jittered Points",
    x = "Species",
    y = "Body Mass (g)"
  )


```

## If time: More Practice with Penguins 

-   Count the number of male and female penguins for each species.
```{r}

penguins %>%
  group_by(species, sex) %>%
  count()
```

-   Find the average penguin mass for each island
```{r}
penguins %>%
  group_by(island) %>%
  summarize(avg_mass = mean(body_mass_g, na.rm = TRUE))

```

-   Find the average bill length by sex
```{r}
penguins %>%
  group_by(sex) %>%
  summarize(avg_bill_length = mean(bill_length_mm, na.rm = TRUE))

```

-   Make a plot of bill_length and bill_depth, but only for the female penguins.
```{r}
ggplot(penguins %>% filter(sex == "female"), aes(x = bill_length_mm, y = bill_depth_mm)) +
  geom_point() +
  labs(
    title = "Bill Length vs Bill Depth (Female Penguins Only)",
    x = "Bill Length (mm)",
    y = "Bill Depth (mm)"
  )

```

-   Make a plot of flipper length and body mass, faceted by island and colored by species.
```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
  geom_point() +
  facet_wrap(~island) +
  labs(
    title = "Flipper Length vs Body Mass",
    subtitle = "Faceted by Island and Colored by Species",
    x = "Flipper Length (mm)",
    y = "Body Mass (g)",
    color = "Species"
  )

```

-   Plot the density of the flipper length, colored by sex
```{r}
ggplot(penguins, aes(x = flipper_length_mm, color = sex)) +
  geom_density() +
  labs(
    title = "Density Plot of Flipper Length",
    x = "Flipper Length (mm)",
    y = "Density",
    color = "Sex"
  )

```

-   Plot the density of the body mass, colored by year

```{r}

ggplot(penguins, aes(x = body_mass_g, color = factor(year))) +
  geom_density() +
  labs(
    title = "Density Plot of Body Mass by Year",
    x = "Body Mass (g)",
    y = "Density",
    color = "Year"
  )


```
